---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# *MorphoRegions*: Analysis of Regionalization Patterns in Serially-Homologous Structures

<!-- badges: start -->
<!-- badges: end -->

*MorphoRegions* is a package package built to computationally identify regions (morphological, functional, etc.) in serially homologous structures such as, but not limited to, the vertebrate backbone. Regions are modeled as segmented linear regressions with each segment corresponding to a region and region boundaries (or breakpoints) corresponding to changes along the serially homologous structure. The optimal number of regions and their breakpoint positions are identified using maximum-likelihood methods without *a priori* assumptions.

This package was first presented in [Gillet et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.03.15.585285v1) and is an updated version of the [*regions* R-package](https://github.com/katrinajones/regions) from [Jones et al. (2018)]({https://www.science.org/doi/abs/10.1126/science.aar3126) with improved computational methods and expanded fitting and plotting options.


## Installing *MorphoRegions*

You can install the released version of *MorphoRegions* from [CRAN](https://CRAN.R-project.org/package=MorphoRegions) with:

```{r eval=F}
install.packages("MorphoRegions")
```

Or the development version from [GitHub](https://github.com/AaGillet/MorphoRegions) with:

```{r eval=F}
# install.packages("remotes")
remotes::install_github("AaGillet/MorphoRegions")
```

## Example

The following example illustrates the basic steps to prepare the data, fit regionalization models, select the best model, and plot the results. See `vignette("MorphoRegions")` or the [*MorphoRegions* website](https://aagillet.github.io/MorphoRegions/) for a detailed guide of the package and its functionalities.


```{r loadpackage}
library(MorphoRegions)
```

#### Preparing the data

Data should be provided as a dataframe where each row is an element of the serially-homologous structure (e.g., a vertebra). One column should contain positional information of each element (e.g., vertebral number) and other columns should contain variables that will be used to calculate regions (e.g., morphological measurements). The `dolphin` dataset contains vertebral measurements of a dolphin with the positional information (vertebral number) in the first column.

```{r loaData}
data("dolphin")
```
```{r printData, echo = F}
knitr::kable(head(dolphin))
```


Prior to analysis, data must be processed into an object usable by `regions` using `process_measurements()`. The `pos` argument is used to specify the name or index of the column containing positional information and the `fillNA` argument allows to fill missing values in the dataset (up to two successive elements).


```{r processData}
dolphin_data <- process_measurements(dolphin, pos = 1)
class(dolphin_data)
```

Data are then ordinated using a Principal Coordinates Analysis (PCO) to reduce dimensionality and allow the combination of a variety of data types. The number of PCOs to retain for analyses can be selected using `PCOselect()` (see the vignette for different methods of PCO axes selection).

```{r PCO}
dolphin_pco <- svdPCO(dolphin_data, metric = "gower")

# Select PCOs with variance > 0.05 :
PCOs <- PCOselect(dolphin_pco, method = "variance",
                  cutoff = .05)
PCOs
```


#### Fitting regressions and selecting the best model


Fitting all possible combinations of segmented linear regressions from 1 region (no breakpoint) to 5 regions (4 breakpoints) along the backbone, with a minimum of 3 vertebrae per region and using a continuous fit (see the vignette for details about fitting options).

```{r calcregions}
regionresults <- calcregions(dolphin_pco, scores = PCOs, noregions = 5,
                             minvert = 3, cont = TRUE, 
                             exhaus = TRUE, verbose = FALSE)
regionresults
```

For each given number of regions, the best fit is selected by minimizing the residual sum of squares (RSS):

```{r modelselect}
models <- modelselect(regionresults)
models
```

The best overall model (best number of regions) is then select by ordering models from best (top row) to worst (last row) using either the AICc or BIC criterion:

```{r modelsupport}
supp <- modelsupport(models)
supp
```

Here, for both criteria, the best model is the 5 regions models with breakpoints at vertebrae 23, 27, 34, and 40. *The breakpoint value corresponds to the last vertebra included in the region, so the first region here is made of vertebrae 8 to 23 included and the second region is made of vertebrae 24 to 27.* The function also returns the **region score**, a continuous value reflecting the level of regionalization while accounting for uncertainty in the best number of regions.


#### Plotting results

Results of the best model (or any other model) can be visualized either as a scatter plot or as a vertebral map.

The **scatter plot** shows the PCO score (here for PCO 1 and 2) of each vertebra along the backbone and the segmented linear regressions (cyan line) of the model to plot. Breakpoints are showed by dotted orange lines.

```{r scatterplot, out.width = "55%", fig.align='center'}
plotsegreg(dolphin_pco, scores = 1:2, modelsupport = supp,
           criterion = "bic", model = 1)
```


In the **vertebral map** plot, each vertebra is represented by a rectangle color-coded according to the region to which it belongs. Vertebrae not included in the analysis (here vertebrae 1 to 7) are represented by grayed rectangles and can be removed using `dropNA = TRUE`.

```{r vertebralmap, fig.show='hold', out.width = "80%", fig.align='center',fig.height=0.65}
plotvertmap(dolphin_pco, name = "Dolphin", modelsupport = supp, 
            criterion = "bic", model = 1)

plotvertmap(dolphin_pco, name = "Dolphin", modelsupport = supp, 
            criterion = "bic", model = 1, dropNA = TRUE)
```


## Citation

To cite `MorphoRegions`, please use: 
```{r citation, eval=F}
citation("MorphoRegions")
```
